---
title: "Survival Analysis - multi-species random effects"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis is following the methods from [Muñoz et al. 2018](http://www.bioone.org/doi/full/10.1642/AUK-17-91.1). In each model, species is included as a random effect and we can therefore extract mean information about the community. Due to the way the estimation works, we can also extract the survival estimates for individual species.

## Data

We are estimating apparent survival for 28 species. The recaptures have been grouped by year, so we have 11 recapture periods (from 33, 3 per year) in order to get annual survival estimates. 

```{r importformat, include = FALSE}
#load packages
library(plyr)
library(tidyverse)
library(data.table)
library(R2jags)
library(cowplot)
library(taxize)
library(knitr)

theme_set(theme_classic() + theme(strip.background = element_blank()))

# specialisation labels
sp_labels <- c("Generalists", "Middle term specialists", "Specialists")

# Load and format banding data ----
data_input <- c("data/banding_sheet.csv", "data/2016_data.csv")

species_names <- read_csv("data/species_names.csv") %>% 
  rename(sci_name1 = `New name`, sci_name2 = `Old name`, common = `English name`) %>% 
  mutate(sci_name2 = toupper(sci_name2))

species_info <- read_csv("data/evaluated_species.csv") %>% 
  inner_join(species_names, by = c("species" = "sci_name2")) %>% 
  #mutate(species = common) %>% 
  select(-sci_name1)
  
dat <- ldply(data_input, fread) %>% 
    # remove juveniles and records with no species or band number
    filter(Age != "Y", 
         !(`Band Number` %in% c("99999", "9999", "", "SIN ANILLO")), 
         !(`Specie Name` %in% c("", "U", "99999", "9"))) %>% 
  # add a habitat and year column
  mutate(habitat = ifelse(Location %in% c("LLAV", "SANA"), "Shrubs", ifelse(Location == "MASE", "Native forest", "Introduced forest")),
         year = str_sub(Session, start = -2) %>% as.factor %>% as.numeric) %>% 
  # select required columns
  select(id = "Band Number", species = "Specie Name", habitat, year) %>% 
  # join to species info to get the specialisms and remove species we are not analysing
  inner_join(species_info) %>% 
  # get unique yearly records (combining across the three sampling sessions)
  group_by(year, id, common, specialization) %>% 
  summarise(habitat = names(which.max(table(habitat)))) %>% 
  mutate(recap = 1) %>% ungroup() %>% 
  rename(species = common)

# create sampsize vector (list of nind for each species) ----
sampsize <- select(dat, species, id) %>% unique %>% group_by(species) %>% summarise(count = n()) %>% pull(count)

# create the y matrix (nind * nocc * nspecies recapture history) ----
y_2d <- select(dat, species, id, year, recap) %>% 
  spread(year, recap, fill = 0)

y_list <- split(y_2d, f = y_2d$species) %>% 
  lapply(., function(x) {
    df = as.matrix(select(x, -species, -id))
    na_rows = matrix(NA, nrow = max(sampsize) - nrow(df), ncol = ncol(df))
    out = rbind(df, na_rows)
  }
  )

y <- simplify2array(y_list)

# create the TSM matrix (nind * nocc * nspecies banding occasion (1 - first, 2 - all others)) ----
tsm_2d <- select(dat, species, id, year) %>% 
  mutate(year = str_pad(year, 2, pad = "0")) %>% 
  # first group-arrange-mutate gets labels the first (1) and all following (2) banding occasions
  group_by(species, id) %>% 
  arrange(species, id, year) %>% 
  mutate(recap = row_number()) %>% 
  mutate(recap = ifelse(recap == 1, 1, 2)) %>% 
  # spread and gather gets the years with no recaps included as 0
  spread(year, recap, fill = 0) %>% 
  gather(key = year, value = recap, -species, -id) %>% 
  # second group-arrange-mutate fills in the blanks
  group_by(species, id) %>% 
  arrange(species, id, year) %>% 
  mutate(recap = ifelse(recap == 0, NA, recap)) %>% 
  fill(recap) %>% 
  spread(year, recap) %>% ungroup

tsm_list <- split(tsm_2d, f = tsm_2d$species) %>% 
  lapply(., function(x) {
    df = as.matrix(select(x, -species, -id))
    na_rows = matrix(NA, nrow = max(sampsize) - nrow(df), ncol = ncol(df))
    out = rbind(df, na_rows)
  }
  )

tsm <- simplify2array(tsm_list)

# create the habitat matrix (nind * nocc * nspecies habitat where recaptured) ----
hab_2d <- mutate(dat, fhabitat = habitat %>% as.factor %>% as.numeric) %>% 
  select(species, id, year, fhabitat) %>% 
  mutate(year = str_pad(year, 2, pad = "0")) %>% 
  # spread and gather gets the years with no recaps included as 0
  spread(year, fhabitat, fill = 0) %>% 
  gather(key = year, value = fhabitat, -species, -id) %>% 
  # second group-arrange-mutate fills in the blanks
  group_by(species, id) %>% 
  arrange(species, id, year) %>% 
  mutate(fhabitat = ifelse(fhabitat == 0, NA, fhabitat)) %>% 
  fill(fhabitat) %>% 
  spread(year, fhabitat) %>% ungroup

hab_list <- split(hab_2d, f = hab_2d$species) %>% 
  lapply(., function(x) {
    df = as.matrix(select(x, -species, -id))
    na_rows = matrix(NA, nrow = max(sampsize) - nrow(df), ncol = ncol(df))
    out = rbind(df, na_rows)
  }
  )

hab <- simplify2array(hab_list)

# create f matrix (nind * nspecies occasion when banded) ----
f <- group_by(dat, species, id) %>% summarise(year = min(year)) %>% 
  group_by(species) %>% mutate(id = row_number()) %>% 
  spread(species, year) %>% select(-id) %>% as.matrix()

# create vector of specialism group (list of specialism for each species) ----
group <- arrange(species_info, species) %>% pull(specialization)


# create vector of family (list of family for each species) ----
fam <- tax_name(query = species_info$species, get = "family", db = "ncbi") %>%
  arrange(query) %>% pull(family) %>% as.factor %>% as.integer

nfam <- max(fam)

# bundle data for jags ----
mod_data <- list(y = y, f = f, tsm = tsm, hab = hab, fam = fam, group = group,
                 nfam = nfam, nind = sampsize, n.occ = dim(y)[2], nspec = dim(y)[3])

# set initial values ---- 
# initial values for z will be set as 1 for each occasion following the occasion
# on which an individual was first banded
zinit <- array(dim = dim(y))
for(s in 1:dim(y)[3]){
  for (i in 1:sampsize[s]) {
    for (t in (f[i,s] + 1):dim(y)[2]) {
      if(f[i,s]+1 <= dim(y)[2]) zinit[i, t, s] <- 1
    }
  }
}

inits <- function(){list(z = zinit)}
params <- c("phi", "p", "speciesphi1", "speciesphi2", "speciesphi1sd", "speciesphi2sd", 
            "speciesp", "mean.phi", "mean.p", "sd.lphi", "sd.lp",
            "indiff", "isdiff", "nsdiff", "beta.group", "allmean.phi", "allsd.phi")
```

## Full model analysis

In the full model, we use a TSM approach (estimate survival between banding and first recapture separately from all other recaptures) to estimate mean apparent survival for each habitat type. We also extract the mean species level estimates, and the mean specialism group level estimates. We estimate p separately for each habitat to control for dectability differences in the three habitats. 

```{r habp_model, eval = FALSE}
mod_out <- jags(data = mod_data, model.file = "code/multi_sp_survival.bug.R", inits = inits, 
                n.iter = 20000, n.chains = 3, parameters.to.save = params)
save(mod_out, file="results/multi_sp_model.rda")
```


```{r analysis}
load("results/multi_sp_model.rda")
sims <- mod_out$BUGSoutput$sims.array
# keep param names to reattach
params <- names(sims[1,1,])
# convert to 2d array
dim(sims) <- c(prod(dim(sims)[1:2]), dim(sims)[3])
colnames(sims) <- params

sims_narrow <- as.tibble(sims) %>% 
  gather(key = param, value = value) %>% 
  mutate(param = str_replace(param, "\\[", "_"),
         param = str_replace(param, "\\]", "")) %>% 
  separate(param, into = c("param_name", "param_index"), sep = "_") 

summ <- as_tibble(mod_out$BUGSoutput$summary, rownames = "param") %>% 
  mutate(param = str_replace(param, "\\[", "_"),
         param = str_replace(param, "\\]", "")) %>% 
  separate(param, into = c("param_name", "param_index"), sep = "_")
```

### Posterior distributions of survival estimates in each habitat type
```{r}
hab_comp <- filter(sims_narrow, param_name == "mean.phi") %>% 
  separate(param_index, into = c("a", "h")) %>% 
  filter(a == 2) %>% 
  mutate(Habitat = factor(h, levels=c(2, 1, 3), 
                          labels=c("Native forest", "Introduced forest", "Shrubs"))) %>% 
  select(Habitat, value)

ggplot(hab_comp, aes(x=value, fill=Habitat)) + 
  geom_density(alpha=0.5, position="identity") + 
  scale_fill_viridis_d() + 
  xlab(expression("Mean "*phi))
```

### Posterior distributions of detection probability in each habitat type
```{r}
hab_p <- filter(sims_narrow, param_name == "mean.p") %>% 
  mutate(Habitat = factor(param_index, levels=c(2, 1, 3), 
                          labels=c("Native forest", "Introduced forest", "Shrubs"))) %>% 
  select(Habitat, value)

hab_p_plot <- ggplot(hab_p, aes(x=value, fill=Habitat)) + 
  geom_density(alpha=0.5, position="identity") + 
  scale_fill_viridis_d() + 
  xlab(expression("Mean "*italic("p"))) + ylab("Density")

hab_p_plot
save_plot("doc/figureS1_habitat_detection_comparison.png", plot = hab_p_plot)
```

Any difference there is between habitats is likely down to the lower detection probability. 

### Posterior distributions of the difference between survival estimates in each habitat type
```{r}
hab_comp <- filter(sims_narrow, param_name %in% c("indiff", "isdiff", "nsdiff")) %>% 
  select(param_name, value)

hab_summary <- filter(summ, param_name %in% c("indiff", "isdiff", "nsdiff")) %>% 
  select(param_name, LCI = `2.5%`, UCI = `97.5%`) %>% gather(CI, value, -param_name)

hab_comp_plot <- ggplot(hab_comp, aes(x=value, fill=param_name)) + 
  geom_density(alpha=0.5, position="identity") + 
  scale_fill_viridis_d(name = "", labels = c(expression(phi[Introduced] - phi[Native]), 
                                  expression(phi[Introduced] - phi[Shrubs]), 
                                  expression(phi[Native] - phi[Shrubs]))) + 
  geom_vline(data = hab_summary, aes(xintercept = value, linetype = param_name)) + 
  scale_linetype_discrete(name = "", labels = c(expression(phi[Introduced] - phi[Native]), 
                                  expression(phi[Introduced] - phi[Shrubs]), 
                                  expression(phi[Native] - phi[Shrubs]))) + 
  xlab(expression("Mean "*phi)) + ylab("Density")

hab_comp_plot
save_plot("doc/figure1_habitat_comparison.png", plot = hab_comp_plot)
```

Note that all 95% CIs of the differences between habitats include zero, so we cannot draw any conclusions about the difference in survival estimate between habitat types. 

### Posterior summaries of group differences in apparent survival
```{r}
group_comp <- filter(sims_narrow, param_name == "speciesphi2") %>% 
  mutate(species_no = param_index %>% as.numeric) %>% 
    inner_join(arrange(species_info, species) %>% 
               mutate(species_no = species %>% as.factor %>% as.numeric)) %>% 
   mutate(specialization = factor(specialization, 
                                  labels=sp_labels))
  

sp_comp_plot <- ggplot(group_comp, aes(x=value, fill=specialization)) + 
  geom_density(alpha=0.5, position="identity") + 
  scale_fill_viridis_d() + 
  xlab(expression("Mean "*phi)) + ylab("Density")

sp_comp_plot

save_plot("doc/figure2_specialism_comparison.png", plot = sp_comp_plot)

beta_group <- filter(summ, param_name == "beta.group") %>% mutate(value = paste0(round(mean, 2), " (", round(`2.5%`, 2), ", ", round(`97.5%`, 2), ")")) %>% pull(value)
```

When we include the group identity as a covariate on phi, while also controlling for phylogenetic correlation by including family as a random factor (following the Muñoz paper), we get a non-important group effect of `r beta_group`. 

### Apparent survival estimates for each species
```{r, fig.height = 8}
survival_comp <- filter(summ, param_name == "speciesphi2") %>% 
  bind_cols(arrange(species_info, species)) %>% 
  select(species = common, specialization, mean_val = mean, LCI = `2.5%`, UCI = `97.5%`) %>% 
  arrange(specialization, species) %>% 
  mutate(species = factor(species, levels = species))

# get the species names in the right order
sp_names <- distinct(survival_comp, species, specialization) %>%
  arrange(specialization, species)

sp_names$specialization <- factor(sp_names$specialization, labels=sp_labels)


# need to know where lines between specializations are
sp_last <- group_by(sp_names, specialization) %>%
  filter(row_number()==n())

sp_last_position <- which(sp_names$species %in% sp_last$species)

sp_first <- group_by(sp_names, specialization) %>%
  filter(row_number()==1)

sp_first_position <- which(sp_names$species %in% sp_first$species)

sp_text_position <- (sp_last_position - sp_first_position)/2 + sp_first_position

specializations <- levels(sp_names$specialization)

pd <- position_dodge(1)

ggplot(survival_comp, aes(x = species, y = mean_val)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = LCI, ymax = UCI)) + 
  geom_vline(xintercept = sp_last_position[1:length(sp_last_position)-1] + 0.5, linetype = "F1") + 
  annotate("text", x = sp_text_position, y = 1, label = specializations) + 
  scale_y_continuous(limits = c(0, 1)) + 
  labs(x = "", y = "Apparent survival") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

write_csv(survival_comp, "results/multi_sp_survival.csv")
```

### Detection probability estimates for each species
```{r, fig.height = 8}
detection_comp <- filter(summ, param_name == "speciesp") %>% 
  bind_cols(arrange(species_info, species)) %>% 
  select(species = common, specialization, mean_val = mean, LCI = `2.5%`, UCI = `97.5%`) %>% 
  arrange(specialization, species) %>% 
  mutate(species = factor(species, levels = species))

pd <- position_dodge(1)

ggplot(detection_comp, aes(x = species, y = mean_val)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = LCI, ymax = UCI)) + 
  geom_vline(xintercept = sp_last_position[1:length(sp_last_position)-1] + 0.5, linetype = "F1") + 
  annotate("text", x = sp_text_position, y = 1, label = specializations) + 
  scale_y_continuous(limits = c(0, 1)) + 
  labs(x = "", y = "Detection probability") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

write_csv(detection_comp, "results/multi_sp_detection.csv")
```

### Variation in apparent survival between habitats for each species 
```{r, fig.height = 8}
habvar_comp <- filter(summ, param_name == "speciesphi2sd") %>% 
  bind_cols(arrange(species_info, species)) %>% 
  select(species = common, specialization, mean_val = mean, LCI = `2.5%`, UCI = `97.5%`) %>% 
  arrange(specialization, species) %>% 
  arrange(specialization, species) %>% 
  mutate(species = factor(species, levels = species))

# get the species names in the right order
sp_names <- distinct(habvar_comp, species, specialization) %>%
  arrange(specialization, species)

sp_names$specialization <- factor(sp_names$specialization, labels=sp_labels)


# need to know where lines between specializations are
sp_last <- group_by(sp_names, specialization) %>%
  filter(row_number()==n())

sp_last_position <- which(sp_names$species %in% sp_last$species)

sp_first <- group_by(sp_names, specialization) %>%
  filter(row_number()==1)

sp_first_position <- which(sp_names$species %in% sp_first$species)

sp_text_position <- (sp_last_position - sp_first_position)/2 + sp_first_position

specializations <- levels(sp_names$specialization)

pd <- position_dodge(1)

ggplot(habvar_comp, aes(x = species, y = mean_val)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = LCI, ymax = UCI)) + 
  geom_vline(xintercept = sp_last_position[1:length(sp_last_position)-1] + 0.5, linetype = "F1") + 
  annotate("text", x = sp_text_position, y = 1, label = specializations) + 
  scale_y_continuous(limits = c(0, 1)) + 
  labs(x = "", y = "Variation in apparent survival across habitats") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

write_csv(habvar_comp, "results/multi_sp_habvar.csv")
```

### Apparent survival estimates for each species by habitat
```{r, fig.width = 20}
survival_hab_comp <- filter(summ, param_name == "phi") %>% 
  separate(param_index, into=c("species", "tsm", "habitat")) %>% 
  filter(tsm == 2) %>% 
  mutate(species_no = species %>% as.integer,
         habitat = factor(habitat, levels=c(2,1,3), labels=c("Native forest", "Introduced forest", "Shrubs"))) %>% 
  select(-species) %>% 
  inner_join(arrange(species_info, species) %>% 
               mutate(species_no = species %>% as.factor %>% as.numeric)) %>% 
  select(species = common, specialization, habitat, mean_val = mean, LCI = `2.5%`, UCI = `97.5%`) %>% 
  arrange(specialization, species) %>%
  mutate(species = factor(species, levels = habvar_comp$species))


pd <- position_dodge(1)

survival_hab_plot <- ggplot(survival_hab_comp, aes(x = species, y = mean_val, group=habitat)) + 
  geom_errorbar(aes(ymin=LCI, ymax=UCI), colour="black", width=.1, position=pd) +
  geom_point(position=pd, aes(shape=habitat), size=3) + 
  geom_vline(xintercept = 1:nrow(sp_names) + 0.5, linetype = "dotted") +
  geom_vline(xintercept = sp_last_position[1:length(sp_last_position)-1] + 0.5, linetype = "F1") + 
  annotate("text", x = sp_text_position, y = 1, label = specializations) + 
  labs(x = "", y = "Apparent Survival", shape = "Habitat") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

save_plot("doc/figureS2_survival_hab_plot.png", survival_hab_plot, base_width = 20, base_height = 10)

write_csv(survival_hab_comp, "results/multi_sp_survival_hab.csv")

survival_hab_plot
```

```{r}
table1 <- filter(summ, param_name %in% c("speciesphi1", "speciesphi2", "speciesp")) %>% 
  mutate(value = paste0(round(mean, 2), " (", round(`2.5%`, 2), ", ", round(`97.5%`, 2), ")")) %>% 
  select(param_name, param_index, value) %>% 
  mutate(param_index = param_index %>% as.integer) %>% 
  spread(param_name, value) %>% 
  bind_cols(arrange(species_info, species)) %>% 
  mutate(specialization = factor(specialization, 
                                 labels=sp_labels)) %>% 
  arrange(specialization, common) %>% 
  select(species = common, specialization, p = speciesp, phi1 = speciesphi1, phi2 = speciesphi2)

write_csv(table1, "doc/table1_survival_estimates.csv")

kable(table1)
```

```{r}
detect_hab_comp <- filter(summ, param_name == "p") %>% 
  separate(param_index, into=c("species", "habitat")) %>% 
  mutate(species_no = species %>% as.integer,
         habitat = factor(habitat, levels=c(2,1,3), labels=c("Native forest", "Introduced forest", "Shrubs"))) %>% 
  select(-species) %>% 
  inner_join(arrange(species_info, species) %>% 
               mutate(species_no = species %>% as.factor %>% as.numeric)) %>% 
  select(species = common, specialization, habitat, mean_val = mean, LCI = `2.5%`, UCI = `97.5%`) %>% 
  arrange(specialization, species) %>%
  mutate(species = factor(species, levels = habvar_comp$species))

table_s2 <- filter(summ, param_name %in% c("phi", "p")) %>% 
  mutate(value = paste0(round(mean, 2), " (", round(`2.5%`, 2), ", ", round(`97.5%`, 2), ")")) %>% 
  separate(param_index, into=c("species", "tsm", "habitat")) %>% 
  filter(tsm == 2) %>% 
  mutate(species_no = species %>% as.integer,
         habitat = factor(habitat, levels=c(2,1,3), labels=c("Native forest", "Introduced forest", "Shrubs"))) %>% 
  select(-species) %>% 
  inner_join(arrange(species_info, species) %>% 
               mutate(species_no = species %>% as.factor %>% as.numeric)) %>% 
  select(species = common, specialization, habitat, value, param_name) %>% 
  spread(param_name, value) %>% mutate(specialization = factor(specialization, 
                                 labels=sp_labels)) %>% 
  arrange(specialization, species)

table_s2 <- bind_rows(survival_hab_comp %>% mutate(param = "phi"),
                      detect_hab_comp %>% mutate(param = "p")) %>% 
  mutate(value = paste0(round(mean_val, 2), " (", round(LCI, 2), ", ", round(UCI, 2), ")")) %>% 
  select(species, specialization, habitat, param, value) %>% 
  spread(param, value) %>% 
  mutate(specialization = factor(specialization, 
                                 labels=sp_labels)) %>% 
  arrange(specialization, species)

write_csv(table_s2, "doc/tableS2_habitat_survival_estimates.csv")

kable(table_s2)
```

